Sensor Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems

### 2023-05-02 latex/intro.tex
Tryin to fix refs

\section{Introduction}\label{section:OED intro}
Experimental design is an important part of many scientific
investigations. When considering an inverse problem, one can often
specify sensor locations (e.g.\ in electricat impedance tomography
\cite{horesh2010impedance}), certain wavelengths (e.g.\ in MRI
\cite{horesh2008mri}), or location of sensors for waves reflecting
from the ground (e.g.\ searching for oil
\cite{horesh2008borehole}). Whatever the allowed set of measurements
is, one should select the optimal measurements to take, in order to
increase accuracy, reduce costs, or both. A set of measurements is
called a \emph{design}, and an \emph{optimal design} is found by
optimizing some \emph{design criterion}. The most popular desgin
criteria are A- and D-optimality

@@

There is a consensus among researchers that sensor clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse} and several remedies to sensor clusterization
were suggeted through the years. In a \emph{clusterization-free
design}, sensor locations are forced to be far from each other
\cite{Ucinski05}. In a frequentist and finite-dimensional setting,
senesor clusterization is avoided by either imposing distance
constraints between measurements, or by introducing correlated errors
which account for both observation error and model error
\cite{Ucinski05}. Merging close measurements is also possible
\cite{fedorov2012}, but this practice ignores sensor clusterization
altogether. In time-series analysis for pharmacokinetic experiments,
sensor clusterization is mitigated by modeling auto-correlation time
in the noise \cite{hooker2009}, and this approach is similar in spirit
to assuming errors are correlatied as in \cite{Ucinski05}.

@@

Many scientists bypass the sensor clusterization problem by
restricting measurements to a coarse grid of the computational domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$ \cite{koval2020,
  alexanderian2021, attia2020, alexanderian2014sparsification,
  AlexanderianPetraStadlerEtAl16, alexanderian2018efficient}, but this
practice requires solvilg a hard combinatorial optimization problem
for choosing optimal sensor locations from a finite set. A solution
typically involves relaxing the problem to that of choosing optimal
measurement ``weight'' in $[0,1]$ for each potential sensor
location. Some authors then add a sparsifying $\ell_1$ penalty term to
the design criterion, which is subsequently thresholded
\cite{horesh2008borehole}, thus resulting in a sparse binary
solution. Others successively relax the $\ell_1$ penalty to a $\ell_0$
penalty via a continuation method
\cite{AlexanderianPetraStadlerEtAl16,
  alexanderian2014sparsification}. A clever alternative is to cast the
problem of finding optimal sensor weights as a stochastic optimization
problem, which in turn is solved by stochastic gradient descent
\cite{attia2022stochastic}. While all of the above mentioned methods
are valid solutions to the prohibitive computational cost of finding
an binary optimal design restricted to a coarse grid, none treats the
elephant in the room: restricting sensors to a coarse grid in $\Omega
\subseteq \mathbb{R}^d, d=1,2,3$ necessarily gives rise to a
sub-optimal design.

Avoiding sensor clusterization is pragmatical: we feel sensor
clusterization is bad (but do not know why), and try to avoid
it. However, all of the abovementioned techniques that avoid sensor
clusterization do so by somehow restricting sensor locations, thus
fundamentaly altering the optimal design problem. To the best of my
knowledge, no study tried to answer the following seemingly trivial
and basic questions: why do sensor clusterization occurs? Is sensor
clusterization really bad? Can we find optimal designs BLABLA?

@@

As a byproduct, the Lagrange multipliers problem for no model error
results in a characterization of D-optimal designs. The key idea is
that a D-optimal design reduces uncertainties where they are highest
first. This is the conclusion of part (5) of Theorem \ref{thm:char},
which we state below and prove in Section \ref{section:vanishing}.

@@

Another byproduct of the proof of Theorem \ref{thm:char} is a novel
linear algebra lemma: We show how to decompose a symmetric positive
definite matrix $M$ as $M = AA^t$, where $A$ has unit norm columns
(Lemma \ref{lemma:lax}). We also generalize two lemmas from linear
algebra to infinite-dimensional settings: In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces, and in Lemma
\ref{lemma:lax} we generalize a lemma \cite{Lax07} for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function.

\subsection{Limitations}\label{subsec:limitations}
The study presented here is not free from drawbacks. First, our
relaxed model does not correspond to any real-life problem. It is
generic enough to be analytically tractable, but one may argue it is
to far removed from any real application. To these claims I would
argue that Mathematicians have a long history of studying models that
are bare-bones simplifications of real systems, e.g. the Ising model
\cite{cipra1987} the Lorenz system \cite{brin}, and the SIR model
\cite{keeling2008}, to name a few. The second drawback of our model is
that we do not show rigorously that clusterization necessarily occurs,
we only show that it does not pose any obstruction to D-optimality.

>>>>

\section{Introduction}\label{section:OED intro}
Measurements of physical processes are indespensable when considering
an inverse problem, choosing an optimal set of measurements is of
utmost importance. In applications, one may have to specify electrode
locations on the skin (e.g.\ in electricat impedance tomography
\cite{horesh2010impedance}), certain wavelengths (e.g.\ in MRI
\cite{horesh2008mri}), or location of sensors for waves reflecting
from the ground (e.g.\ searching for oil
\cite{horesh2008borehole}). Whatever the allowed set of measurements
is, one should select the optimal measurements to take, in order to
increase accuracy, reduce costs, or both. A set of measurements is
called a \emph{design}, and an \emph{optimal design} is found by
optimizing some \emph{design criterion}. Two of the most popular
desgin criteria are A- and D-optimality (see \cite{Chaloner1995} for a
thorough review).

@@

There is a consensus among researchers that sensor clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse} and several remedies to sensor clusterization
were suggeted through the years. Merging ``close'' measurements is
possible, \cite{fedorov2012}, but this practice ignores sensor
clusterization altogether. In a \emph{clusterization-free design},
sensor locations are forced to be far from each other
\cite{Ucinski05}. In a frequentist and finite-dimensional setting,
senesor clusterization is avoided by either imposing distance
constraints between measurements, or by introducing correlated errors
which account for both observation error and model error
\cite{Ucinski05}. In time-series analysis for pharmacokinetic
experiments, sensor clusterization is mitigated by modeling
auto-correlation time in the noise \cite{hooker2009}, and this
approach is similar in spirit to assuming errors are correlatied
\cite{Ucinski05}.

@@

In spatial problems, where sensors are placed in some domain $\Omega
\subseteq \mathbb{R}^d, d=1,2,3$, many scientists bypass the sensor
clusterization problem by restricting measurements to a coarse grid in
$\Omega$ \cite{koval2020, alexanderian2021, attia2020,
  alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. Unfortunately, this practice requires
solvilg a hard combinatorial optimization problem for choosing optimal
sensor locations from a finite set. A solution typically involves
relaxing the problem to choosing optimal measurement ``weight'' in
$[0,1]$ for each potential sensor location. Some authors then add a
sparsifying $\ell_1$ penalty term to the design criterion, which is
subsequently thresholded \cite{horesh2008borehole}, thus resulting in
the desired binary design. Others successively relax the $\ell_1$
penalty to a $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal sensor weights as a stochastic
optimization problem, which in turn is solved by stochastic gradient
descent \cite{attia2022stochastic}????. While all of the above
mentioned methods are valid solutions to the prohibitive computational
cost of finding a binary optimal design restricted to a coarse grid,
none treats the elephant in the room: restricting sensor locations to
a coarse grid in $\Omega \subseteq \mathbb{R}^d, d=1,2,3$ necessarily
gives rise to a sub-optimal design.

Avoiding sensor clusterization is pragmatical: we feel sensor
clusterization is bad (but do not know why), and try to avoid
it. However, all of the abovementioned techniques that avoid sensor
clusterization do so by somehow restricting sensor locations, thus
fundamentaly altering the optimal design problem. To the best of my
knowledge, no study tried to answer the following seemingly trivial
and basic questions: why do sensor clusterization occurs? Should
sensor clusterization really be avoided? How imposing a correlation
structure between obsevations mitigates sensor clusterization? Can we
find truly optimal designs that do not exhibit sensor clusterization?

@@

the Lagrange multipliers problem for vanishing model error results in
a characterization of D-optimal designs. The key idea is that a
D-optimal design reduces uncertainties where they are highest
first. This is the conclusion of part (5) of Theorem \ref{thm:char},
stated below and proved in Section \ref{section:vanishing}.

@@

An interesting byproduct of the proof of Theorem \ref{thm:char} is a
novel lemma in linear algebra: For a symmetric positive definite
matrix $M \in \mathbb{R}^{k \times k}$ with $\ttr M = m \in
\mathbb{N}$, we decompose $M = AA^t$, where $A$ has unit norm columns
(Lemma \ref{lemma:free}). We also generalize two lemmas from linear
algebra to infinite-dimensional settings, see Lemmas \ref{lemma:MDL}
and \ref{lemma:lax}.

%%  we prove a matrix determinant lemma in Hilbert spaces, and in
%%  Lemma \ref{lemma:lax} we generalize a lemma for calculating
%%  $\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
%%  operator valued function \cite{Lax07}.

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. It is generic enough to
be analytically tractable, but one may argue it is too far removed
from any real application. To these claims I would argue that
Mathematicians have a long history of studying models that are
bare-bones simplifications of real systems. E.g. the Ising model
\cite{cipra1987}, the Lorenz system \cite{brin}, and the SIR model
\cite{keeling2008}, to name a few.

%% The second drawback of our model is that we do not show rigorously
%% that clusterization necessarily occurs, we only show that it does
%% not pose any obstruction to D-optimality.

### 2023-05-02 latex/prelims.tex
Tryin to fix refs

\subsection{Overview}
In this section I will present notations that will be used throughout
this article. These notations are summarized in Section
\ref{subsec:notation}. The theoretical foundations for inverse
problems over function spaces can be found in
\cite{Stuart10}.

@@

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. Thus, the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}. Bayesian D-optimal designs for
inverse problems on function spaces were first studied by
\cite{AlexanderianGloorGhattas14, AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}, and just like in finite dimensions,
a D-optimal design maximizes the Kullback-Leibler divergence between
posterior and prior measures.
%% For a linear inverse problem with Gaussian prior and Gaussian
%% noise, a D-optimal design minimizes an expression analogous to the
%% determinant of the posterior covariance operator
%% \cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
%% below).
D-optimal designs in infinite-dimensional
Hilbert spaces were investigated in
\cite{AlexanderianGloorGhattas14}. Their main result is summarized, in
the notation presented above, in the following theorem\footnote{In
\cite{AlexanderianGloorGhattas14, alexanderian2018efficient}, results
are stated for $\Sigma=I$ (implied by $\modcov = 0,\sigma^2=1$), but
the authors also mention these results hold for more general
covariance matrices \cite[p. 681]{AlexanderianGloorGhattas14}.}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}

>>>>

%% \subsection{Overview}
In this section I will present notations that will be used throughout
this article. These notations are summarized in Section
\ref{subsec:notation}. The theoretical foundations for inverse
problems over function spaces can be found in \cite{Stuart10}.

@@

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. Thus, the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}. Bayesian D-optimal designs for
inverse problems on function spaces were first studied by
\cite{AlexanderianGloorGhattas14, alexanderian2018efficient}, and just like in finite dimensions,
a D-optimal design maximizes the Kullback-Leibler divergence between
posterior and prior measures.
%% For a linear inverse problem with Gaussian prior and Gaussian
%% noise, a D-optimal design minimizes an expression analogous to the
%% determinant of the posterior covariance operator
%% \cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
%% below).
D-optimal designs in infinite-dimensional
Hilbert spaces were investigated in
\cite{AlexanderianGloorGhattas14}. Their main result is summarized, in
the notation presented above, in the following theorem\footnote{In
\cite{AlexanderianGloorGhattas14, alexanderian2018efficient}, results
are stated for $\Sigma=I$ (implied by $\modcov = 0,\sigma^2=1$), but
the authors also mention these results hold for more general
covariance matrices \cite[p. 681]{AlexanderianGloorGhattas14}.}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}
